The onset of hydrodynamically driven waving of marine vegetation in flow

### 2014-09-26 Paper/grass.tex
Continue to improve language

\graphicspath{ {images/}{images/Growth_Rate_Cont_Plots/} }
\newcommand{\bn}{{\boldsymbol{\hat{n}}}}
\newcommand{\bt}{{\boldsymbol{\hat{t}}}}
\newcommand{\bu}{\mathbf{u}}
\newcommand{\grad}{\mathbf{\nabla}}
\newcommand{\del}{\partial}
\newcommand{\hg}{h_g}
\newcommand{\Rey}{\text{Re}}
\newcommand{\Ndg}{\tilde{N}_g}
\newcommand{\monami}{\textit{monami }}
% \newcommand{\shreyas}[1]{{\bf (#1)}}
\newcommand{\shreyas}[1]{}
\definecolor{tableShade}{gray}{0.8}

\begin{document}
\title{The onset of hydrodynamically driven waving of marine grass in flow}
\author{Ravi Singh}
\affiliation{Brown University, Providence RI 02912 USA}
\author{L. Mahadevan}
\affiliation{Harvard University, Cambridge MA 02138 USA}
\author{M. M. Bandi}
\affiliation{Collective Interactions Unit, OIST Graduate University, 1919-1 Tancha, Onna-son, Okinawa, Japan 904-0495}
\author{Amala Mahadevan}
\affiliation{Woods Hole Oceanographic Institute, Woods Hole MA USA}
\author{Shreyas Mandre}
\affiliation{Brown University, Providence RI 02912 USA}

@@

We develop a self-consistent model of the flow past rigid vegetation, where the vegetation is represented by a continuum drag field. To explain the onset of spontaneous waving, we perform a linear stability analysis of the steady flow resulting from this model. We show that a threshold flow condition for the spontaneous waving of marine grass arises naturally in this framework due to the drag exerted by the vegetation on the flow. Beyond the threshold, we find two unstable modes; first is localized in a boundary layer near the canopy top, and second which spans the whole water column. The first mode shares many common features with the Kelvin-Helmholtz instability, but is demonstrated to be different from it. The predictions of the model for threshold condition and the resulting dominant frequency of waving agree with laboratory observations. 

@@

The simplest response of these flexible grass canopies to steady currents is the formation of coherent large amplitude oscillations, known as \monami \cite{AckermanOkubo93}.  

@@

Current explanations of \monami invoke the existence of a shear layer near the top of the canopy  \cite{Ghisal02,Raupach96} and its instability through a mechanism similar to the Kelvin-Helmholtz instability. 
This instability is thought to lead to coherent eddies over the sea grass canopy which then drives large amplitude synchronous oscillations.

@@

While the shear layer model appears to predict the frequency of \monami consistent with experimental observations, several aspects of existing theory remain unsatisfactory. 

@@

(ii) Classical free shear flow is known to be unstable for all Reynolds numbers \cite{drazin}, whereas observations in the lab \cite{Ghisal02} and the field \cite{Grizzle96} indicate the existence of a threshold 
flow speed below which \monami is not observed. 
(iii)The thickness of the free shear layer, defined as the momentum thickness of the boundary layer near the grass tip, is in many cases comparable to the unvegetated layer thickness,
and therefore inconsistent with the assumptions of the free shear layer instability.
Here we present a self-consistent mathematical model that accounts correctly for these effects 
while also explaining lab experiments and field observations.

@@

The drag exerted by the grass bed on the flow is central to the hypothesized mechanism \cite{Ghisal02} of the instability underlying \monami because it realizes the shear layer structure of the flow. 

@@

Indeed, a more detailed calculation undertaken for terrestrial grass demonstrated \cite{Delangre06} that when the timescale of the instability is well-separated from the period of natural oscillations of the grass bed, 
the flow instability is similar to the instability of a shear layer. 

@@

We use a mean field model for the coupling between the flow and the canopy. 
The force exerted by the grass is approximated via a continuous body force $\mathbf{f}=-N_g\mathbf{f_d}$,  
where $\mathbf{f_{d}}$ is drag force per unit length of the grass blade, and $N_g$ the grass number density per unit area \cite{Vivoni98,Nepf99,Ghisal02,Delangre04,Delangre06}. 
In the canopy the equations of continuity and momentum balance can be written as 

@@

\quad \grad\cdot\bu = 0,\hspace{3mm} \rho \left(\bu_{t}+\bu.\grad\bu \right) = -\grad P+\mu\grad^{2}\bu +\mathbf{f}+\rho\mathbf{g}

@@

$P$ the pressure, $\mu$ the eddy viscosity (the flow is usually turbulent) and $\mathbf{g}$ the acceleration due to gravity. 
The Reynolds number of the flow based on the scale of a grass blade is expected to be $O(10^2-10^3)$, and hence the drag force is dominated by pressure and modeled as 
$\mathbf{f_{d}}=C_N \rho u_{N}^{2}d\bn + C_{T}\rho u_{T}^{2}d\bt$ \shreyas{citation needed} 
where $d$ is average width of grass blade $C_{N}$ and $C_{T}$ are the normal and tangential drag coefficients respectively; $u_{T}$, $u_{N}$ are velocity components along and
normal to grass while $\bt,\bn$ are unit vectors along and normal to grass. 
We expect $C_T \ll C_N$ and take $C_T=0$ for rest of analysis. 
In the field, both $C_N$ and $N_g$ vary with distance from the bottom but we do not expect these variations to be central to the mechanism and therefore take $C_N$ and $N_g$ to be constants. 

@@

Such a steady state flow driven by constant pressure gradient satisfies 

@@

In the limit of large grass density, shear stress exerted by the bottom surface is estimated to be negligible in comparison with the vegetation drag and hence the bottom surface ($y=0$) is approximated by zero shear.   

@@

Flow within grass has approximately constant velocity $U_g \sim \sqrt{\frac{dP/dx}{\rho C_N dN_g}}$ arising from the balance between drag and pressure gradient, and the flow outside has a simple parabolic velocity profile due to the balance between viscous forces and the pressure gradient. 
Continuity of flow speed and shear stresses at canopy top gives rise to a boundary layer near the tip of the grass.

@@

 and it's approximation with $U_0=7.28$ cm/s and $\delta = 5.02$ cm with our model. The canopy extends upto $y=h_g$ in the water column of depth $2H$. 
The steady velocity profile can be decomposed into three regions, (i) A parabolic profile in the unvegetated region, (ii) a uniform profile deep within the grass bed, and (iii) a boundary layer of thickness $\delta$ near the canopy top. 

@@

This boundary layer near the canopy top results from purely local dynamics independent of the influence of the boundaries, and we identify it as a free shear layer \cite{Ghisal02} in the previous explanation of \monami. 
In our model the balance between viscous forces and the drag force in the shear layer $(\mu u_\text{sl}/\delta^2 \sim \rho C_N d N_g u_\text{sl}^2)$ and the continuity of shear stress across the grass tip $(u_\text{sl}/\delta \sim U_0/H)$, leads to a shear layer thickness $\delta=(\Rey\tilde{N_g})^{-1/3}H$, where $\Ndg = \left(C_N d H N_g\right)$ is the vegetation frontal area per bed area, $\Rey=\rho U_0 H/\mu$ is the Reynolds number of the flow, $u_{sl}$ is the velocity scale in shear layer and $U_0 = {(dP/dx)~H^2}/{\mu}$ is the velocity scale in the unvegetated region. 
This dependence of the boundary layer thickness on the grass density $N_g$, shown in fig.~\ref{basicflow}, gives us a way to systematically investigate its effect on the instability mechanism.

@@

We are left with three non-dimensional parameters in the model, ie. $\Rey$, $\Ndg$, and the submergence ratio of the grass $h_g/H$. 
We also use $\delta/H$ in lieu of $\Ndg$ to parametrize the grass density and help elucidate the mechanism of the instability. 

@@

We solve this eigenvalue problem and a threshold in $\Rey$ of $O(100)$, above which the flow is unstable (real part of $\sigma$ is positive), emerges from the 
solution of eq.~\eqref{Orr-somerfield}. The dependence of this critical $\Rey$, and the corresponding fastest growing wavenumber $k$, on $\delta/H$ and $\hg/H$ is shown 
in fig.~\ref{Re_vs_delta}, and is found to compare well with experimental observations \cite{Ghisal02}.
The threshold Reynolds number increases with the grass density, indicating a competition between the 
destabilizing shear in the flow, and the stabilizing effect of damping due to vegetation drag.
Previous linear stability calculations for terrestrial grass do not exhibit a threshold flow condition 
either due to the exclusion of the vegetation drag in their model \cite{Raupach96}, or because of ad-hoc
assumptions for the mean velocity profile \cite{Raupach96,Delangre06}.

\begin{figure*}
\begin{tabular}{cccc}
{\includegraphics[scale = 0.67]{Set4_dens28_imgsc}} &
{\includegraphics[scale = 0.67]{Set4_dens30_imgsc}} &
{\includegraphics[scale = 0.67]{Set4_dens32_imgsc}} &
{\includegraphics[scale = 0.67]{Set4_dens34_imgsc}} \\
{\includegraphics[scale = 0.67]{Set5_dens38_imgsc}} &
{\includegraphics[scale = 0.67]{Set5_dens40_imgsc}} &
{\includegraphics[scale = 0.67]{Set5_dens42_imgsc}} &
{\includegraphics[scale = 0.67]{Set5_dens46_imgsc}} \\
%{\includegraphics[scale = 0.67]{Set6_dens32_imgsc}} &
%{\includegraphics[scale = 0.67]{Set6_dens34_imgsc}} &
%{\includegraphics[scale = 0.67]{Set6_dens36_imgsc}} &
%{\includegraphics[scale = 0.67]{Set6_dens38_imgsc}}
\end{tabular}
\caption{$Re(\sigma)$ as function of wavenumber and Reynolds number for a specified grass number density $\Ndg$ along with the neutral curve ($Re(\sigma)$=0), for parameters shown in the corresponding panel.  
As $\Ndg \propto N_g$ increases, the unstable region splits into two labeled as ``Mode 1'' and ``Mode 2''. 
For $\Ndg$ below a critical value, the Mode 1 sets the threshold $\Rey$ whereas above the critical $\Ndg$ onset is determined by Mode 2.}
\label{K_Re_sigma_set3}
\end{figure*}

Comparison of frequency (Im$(\sigma)$) of the fastest growing mode with that of experimental observations of frequencies in
the lab scale experiments, for cases where the vegetation was sufficiently dense to be modeled by a continuum drag field, is also shown in the inset in fig.~\ref{Re_vs_delta}b. 
The observed frequencies are associated with the peaks in velocity spectra, frequency of $\monami$ and frequency of 
vortex passage \cite{Ghisal02}, and compare well with the predicted frequencies. 
%Regardless of the assumptions, the introduction of vegetation drag is observed to damp the linear modes\cite{Delangre06}.

\begin{table*}
\rowcolors{3}{tableShade}{white}  %% start alternating shades from 3rd row
\renewcommand{\arraystretch}{1.4}
 \begin{tabular}{l|c|c|c}
			& Kelvin-Helmholtz 				& Mode 1 		& Mode 2 \\ \hline
 Base velocity profile 	& $U(y) = U_0 \tanh(y/\delta)$			& \multicolumn{2}{c}{$-{dP}/{dx}+\mu U''(y) +S(y) \rho C_N d N_gU^2=0$} \\
 Domain 		& $-\infty < y < \infty$			& \multicolumn{2}{c}{$-1<y<1$} \\
 Inflection point	& exists at $y=0$				& \multicolumn{2}{c}{$U''(y)$ discontinuous at $y=\hg$} \\
 Shear layer thickness	& $\delta$					& \multicolumn{2}{c}{$\delta \sim  H\left(\Rey \Ndg \right)^{-1/3}$} \\
 Linearized dynamics	& $\left(\sigma+ikU\right) \left(D^2-k^2\right)\phi =  ikU_{yy}\phi$		& \multicolumn{2}{c}{Equation \eqref{Orr-somerfield}} \\
 Critical parameters	& none						& $\Rey \propto \Ndg^{1/2}$ 	& $\Rey \propto \Ndg$ \\
 Most unstable $k$ as $\delta \to 0$	& $\propto \delta^{-1}$		& $\propto \delta^{-1}$	& $O(1)$ \\
 Mode localized?	& yes, near $y=0$				& ~~~~yes, near $y=\hg$~~~~			& no
 \end{tabular}
 \caption{Comparison between Kelvin-Helmholtz instability and the two modes resulting from solution of \ref{Orr-somerfield}.}
 \label{tab:comparison}
\end{table*}
To better understand the mechanism of waving and the origin of the threshold, we consider the behavior of the instability as a function of the vegetation drag characterized by $\Ndg$.
The fastest growing wavenumber first increases proportional to $H/\delta$, but at a critical grass density discontinuously jumps and remains $O(1)$ (see fig.~\ref{Re_vs_delta}). 
To aid in explaining the behavior, we show heat maps of Re$(\sigma)$ as a function of $\Rey$ and $k$, for different $\hg/H$ and $\Ndg$ in fig.~\ref{K_Re_sigma_set3}. 
The lowest $\Rey$ on the neutral curve sets the threshold. 
We observe that as $\Ndg$ increases, the unstable region splits into two. We refer to the unstable region with the higher $k$ as ``Mode 1'', and the one with the lower $k$ as ``Mode 2''. 
The unstable region for Mode 1, depending on $\hg/H$ either recedes to higher $\Rey$ or shrinks to zero size, as the grass density increases, causing the most unstable mode to jump.

The distinct asymptotic behavior of the two modes as $\Ndg \gg 1$ allows us to understand the mechanism of the instability. 
Mode 1 asymptotically localizes to the boundary layer near the grass tip, and exhibits an asymptotic behavior with $k \sim O(H/\delta)$, and $\Rey \sim (H/\delta)$ (or $\Rey \propto {\Ndg}^{1/2}$) at the threshold. 
This limit can be understood through asymptotically estimating the sizes of the terms in eq.~\eqref{Orr-somerfield} using $D\sim \delta^{-1}$, $\sigma \sim \delta^{-1}$, and $U\sim \delta$; the magnitude of the advection term is $\Rey/\delta^3 \sim \Rey^2 \Ndg $, and the viscous term (or the vegetation drag term, which is found to scale identically) in the boundary layer is $\delta^{-4} \sim (\Rey \Ndg)^{4/3}$. 
The terms balance when $\Rey \sim {\Ndg}^{1/2}$ (or $\Rey \sim H/\delta$) leading to a simplification of eq.~\eqref{Orr-somerfield} as
\begin{align}
(D^2-k^2)^2 \phi- 2 \Rey \Ndg D( SUD\phi) = \sigma \Rey (D^2-k^2)\phi,
\label{eqn:mode1asymp}
\end{align} 
in a region of thickness O($\delta$) near $y=\hg$.
Since the relative magnitude of the terms on the r.h.s. to those on the l.h.s. of this equation is $\Rey/\Ndg^{1/2}$, we expect identical asymptotic behavior for fixed $\Rey/\Ndg^{1/2}$. Therefore the threshold obtained for Mode 1 is $\Rey \propto \Ndg^{1/2}$ (or $\Rey \propto H/\delta$) explaining the numerically observed asymptote (see fig.~\ref{Re_vs_delta}). 
This analysis also concludes that the mode structure is self-similar over the length scale $\delta$ for fixed $\Rey/\Ndg^{1/2}$; the verification of this expectation is shown in fig.~\ref{Asymptotic_mode}, supporting this argument.

>>>>

\graphicspath{ {images/}{images/Growth_Rate_Cont_Plots/} }
\newcommand{\bn}{{\boldsymbol{\hat{n}}}}
\newcommand{\bt}{{\boldsymbol{\hat{t}}}}
\newcommand{\bu}{\mathbf{u}}
\newcommand{\grad}{\mathbf{\nabla}}
\newcommand{\del}{\partial}
\newcommand{\hg}{h_g}
\newcommand{\Rey}{\text{R}}
\newcommand{\Ndg}{\tilde{N}_g}
\newcommand{\monami}{\textit{monami}}
% \newcommand{\shreyas}[1]{{\bf (#1)}}
\newcommand{\shreyas}[1]{}
\definecolor{tableShade}{gray}{0.8}

\begin{document}
\title{The onset of hydrodynamically driven waving of marine vegetation in flow}
\author{Ravi Singh}
\affiliation{Brown University, Providence RI 02912 USA}
\author{L. Mahadevan}
\affiliation{Harvard University, Cambridge MA 02138 USA}
\author{M. M. Bandi}
\affiliation{OIST Graduate University, Okinawa 904-0495, Japan}
\author{Amala Mahadevan}
\affiliation{Woods Hole Oceanographic Institution, Woods Hole MA 02543 USA}
\author{Shreyas Mandre}
\affiliation{Brown University, Providence RI 02912 USA}

% \begin{abstract}
% We develop a self-consistent model of the flow past rigid vegetation, where the vegetation is represented by a continuum drag field. To explain the onset of spontaneous waving, we perform a linear stability analysis of the steady flow resulting from this model. We show that a threshold flow condition for the spontaneous waving of marine grass arises naturally in this framework due to the drag exerted by the vegetation on the flow. Beyond the threshold, we find two unstable modes; first is localized in a boundary layer near the canopy top, and second which spans the whole water column. The first mode shares many common features with the Kelvin-Helmholtz instability, but is demonstrated to be different from it. The predictions of the model for threshold condition and the resulting dominant frequency of waving agree with laboratory observations. 
% \end{abstract}

@@

We develop a self-consistent mathematical model for the onset of synchronous waving of dense marine vegetation driven by a hydrodynamic instability.
The vegetation, assumed rigid, only exerts a drag force on the flow.
We show that a linear instability of a steady pressure-driven flow explains the experimentally observed flow threshold for waving and the waving frequency, confirming the hydrodynamic mechanism for the onset of waving, where the vegetation flexibility merely serves to visualize the flow.
We also demonstrate that our instability is different from the Kelvin-Helmholtz instability.

@@

The simplest response of these flexible vegetation canopies to steady currents is the formation of coherent large amplitude oscillations, known as \monami ~\cite{AckermanOkubo93}.  

@@

Current explanations of \monami ~invoke the existence of a shear layer due to vegetation drag  \cite{Ghisal02,Raupach96} and its instability through a mechanism similar to the Kelvin-Helmholtz instability. 
This instability is thought to lead to coherent eddies over the sea grass meadow, which then drive large amplitude synchronous oscillations.

@@

While the shear layer model appears to predict the frequency of \monami ~consistent with experimental observations, several aspects of existing theory remain unsatisfactory. 

@@

(ii) Classical free shear flow is known to be unstable for all Reynolds numbers \cite{drazin}, whereas observations in the lab \cite{Ghisal02} and the field \cite{Grizzle96} indicate the existence of a threshold flow speed below which \monami ~is not observed. 
(iii)The thickness of the free shear layer, defined as the momentum thickness of the boundary layer near the top of the meadow, is in many cases comparable to the unvegetated layer thickness, and therefore inconsistent with the assumptions of the free shear layer instability.
Here we present a self-consistent mathematical model for the hydrodynamic linear stability analysis that accounts correctly for these effects, while also explaining lab experiments and field observations.

@@

The drag exerted by the grass bed on the flow is central to the hypothesized mechanism \cite{Ghisal02} of the instability underlying \monami ~because it realizes the shear layer structure of the flow. 

@@

Indeed, a more detailed calculation modeling terrestrial grass demonstrated \cite{Delangre06} that when the timescale of the instability is well-separated from the period of natural oscillations of the grass bed, the flow instability is similar to the instability of a shear layer. 

@@

We use a mean field model for the coupling between the flow and the vegetation. 
The force exerted by the vegetation is approximated via a continuous body force $\mathbf{f}=-N_g\mathbf{f_d}$,  
where $\mathbf{f_{d}}$ is drag force per unit length of the grass blade, and $N_g$ the grass blade number density per unit area \cite{Vivoni98,Nepf99,Ghisal02,Delangre04,Delangre06}. 
In the vegetation, the equations of continuity and momentum balance can be written as 

@@

\quad \grad\cdot\bu = 0,\hspace{3mm} \rho \left(\bu_{t}+\bu.\grad\bu \right) = -\grad p+\mu\grad^{2}\bu +\mathbf{f}+\rho\mathbf{g}

@@

$p$ the pressure, $\mu$ the eddy viscosity (the flow is turbulent) and $\mathbf{g}$ the acceleration due to gravity. 
The Reynolds number of the flow based on the scale of the grass blade is expected to be $O(10^2-10^3)$, implying that the form drag on the vegetation dominates over the skin-friction. Hence we model it as 
$\mathbf{f_{d}}=C_N \rho u_{N}^{2}d\bn$, \shreyas{citation needed} 
where $d$ is average width of grass blade projected in the flow direction, $C_{N}$ the drag coefficient, $u_{N}$ and $\bn$ are the velocity component and unit vector normal to the grass. 
In the field, both $C_N$ and $N_g$ vary with distance from the bottom but we do not expect these variations to be central to the instability mechanism and therefore take $C_N$ and $N_g$ to be constants. 

@@

Such a steady state flow driven by constant pressure gradient $dP/dx$ satisfies 

@@

In the limit of large vegetation density, shear stress exerted by the bottom surface is expected to be negligible compared to the vegetation drag, and hence the bottom surface ($y=0$) is approximated by zero shear.   

@@

Flow within vegetation has approximately constant velocity $U_g \sim \sqrt{\frac{dP/dx}{\rho C_N dN_g}}$ arising from the balance between drag and pressure gradient, and the flow outside has a simple parabolic velocity profile due to the balance between viscous forces and the pressure gradient. 
Continuity of flow speed and shear stresses at the top of the meadow gives rise to a boundary layer.

@@

 and it's approximation with $U_0=7.28$ cm/s and $\delta = 5.02$ cm in our model. The meadow extends upto $y=h_g$ in the water column of depth $2H$. 
The steady velocity profile can be decomposed into three regions, (i) A parabolic profile in the unvegetated region, (ii) a uniform profile deep within the vegetation, and (iii) a boundary layer of thickness $\delta$ near the top of the meadow. 

@@

This boundary layer results from purely local dynamics, independent of the influence of the boundaries, and we identify it as a free shear layer \cite{Ghisal02} in the previous explanation of \monami. 
In our model the balance between viscous forces and the drag force in the shear layer $(\mu u_\text{sl}/\delta^2 \sim \rho C_N d N_g u_\text{sl}^2)$ and the continuity of shear stress across the top of the vegetation $(u_\text{sl}/\delta \sim U_0/H)$, leads to a shear layer thickness $\delta=(\Rey\tilde{N_g})^{-1/3}H$, where $\Ndg = \left(C_N d H N_g\right)$ is the vegetation frontal area per bed area, $\Rey=\rho U_0 H/\mu$ is the Reynolds number of the flow, $u_{sl}$ is the velocity scale in shear layer and $U_0 = {(dP/dx)~H^2}/{\mu}$ is the velocity scale in the unvegetated region. 
This dependence of the boundary layer thickness on the vegetation density $N_g$, shown in fig.~\ref{basicflow}, gives us a way to systematically investigate its effect on the instability mechanism.

@@

We have three non-dimensional parameters in the model, ie. $\Rey$, $\Ndg$, and the submergence ratio of the vegetation $h_g/H$. 
We also use $\delta/H$ in lieu of $\Ndg$ to parametrize the vegetation density and help elucidate the mechanism of the instability. 

@@

We solve this eigenvalue problem and a threshold in $\Rey$ of $O(100)$, above which the flow is unstable (Re$(\sigma)>0$), emerges from the 
solution of eq.~\eqref{Orr-somerfield}. The dependence of this critical $\Rey$, and the corresponding fastest growing wavenumber $k$, on $\delta/H$ and $\hg/H$ is shown 
in fig.~\ref{Re_vs_delta}, and is found to compare well with experimental observations \cite{Ghisal02}.
The threshold Reynolds number increases with the vegetation density, indicating a competition between the 
destabilizing shear in the flow, and the stabilizing effect of damping due to vegetation drag.
Previous linear stability calculations for terrestrial grass do not exhibit a threshold flow condition 
either due to the exclusion of the vegetation drag in their model \cite{Raupach96}, or because of ad-hoc
assumptions for the mean velocity profile \cite{Raupach96,Delangre06}.

\begin{figure*}
\begin{tabular}{cccc}
{\includegraphics[scale = 0.67]{Set4_dens28_imgsc}} &
{\includegraphics[scale = 0.67]{Set4_dens30_imgsc}} &
{\includegraphics[scale = 0.67]{Set4_dens32_imgsc}} &
{\includegraphics[scale = 0.67]{Set4_dens34_imgsc}} \\
{\includegraphics[scale = 0.67]{Set5_dens38_imgsc}} &
{\includegraphics[scale = 0.67]{Set5_dens40_imgsc}} &
{\includegraphics[scale = 0.67]{Set5_dens42_imgsc}} &
{\includegraphics[scale = 0.67]{Set5_dens46_imgsc}} \\
%{\includegraphics[scale = 0.67]{Set6_dens32_imgsc}} &
%{\includegraphics[scale = 0.67]{Set6_dens34_imgsc}} &
%{\includegraphics[scale = 0.67]{Set6_dens36_imgsc}} &
%{\includegraphics[scale = 0.67]{Set6_dens38_imgsc}}
\end{tabular}
\caption{$Re(\sigma)$ as function of wavenumber and Reynolds number for a specified vegetation number density $\Ndg$ along with the neutral curve ($Re(\sigma)$=0), for parameters shown in the corresponding panel.  
As $\Ndg \propto N_g$ increases, the unstable region splits into two labeled as ``Mode 1'' and ``Mode 2''. 
For $\Ndg$ below a critical value, the Mode 1 sets the threshold $\Rey$ whereas above the critical $\Ndg$ onset is determined by Mode 2.}
\label{K_Re_sigma_set3}
\end{figure*}

Comparison of frequency (Im$(\sigma)$) of the fastest growing mode with that of experimental observations of frequencies in
the lab scale experiments, for cases where the vegetation was sufficiently dense to be modeled by a continuum drag field, is also shown in the inset in fig.~\ref{Re_vs_delta}b. 
The observed frequencies are associated with the peaks in velocity spectra, frequency of \monami ~and frequency of 
vortex passage \cite{Ghisal02}, and compare well with the predicted frequencies. 
%Regardless of the assumptions, the introduction of vegetation drag is observed to damp the linear modes\cite{Delangre06}.

\begin{table*}
\rowcolors{3}{tableShade}{white}  %% start alternating shades from 3rd row
\renewcommand{\arraystretch}{1.4}
 \begin{tabular}{l|c|c|c}
			& Kelvin-Helmholtz 				& Mode 1 		& Mode 2 \\ \hline
 Base velocity profile 	& $U(y) = U_0 \tanh(y/\delta)$			& \multicolumn{2}{c}{Equation \eqref{base_equ}} \\
 Domain 		& $-\infty < y < \infty$			& \multicolumn{2}{c}{$-1<y<1$} \\
 Inflection point	& exists at $y=0$				& \multicolumn{2}{c}{$U''(y)$ discontinuous at $y=\hg$} \\
 Shear layer thickness	& $\delta$					& \multicolumn{2}{c}{$\delta \sim  H\left(\Rey \Ndg \right)^{-1/3}$} \\
 Linearized dynamics	& $\left(\sigma+ikU\right) \left(D^2-k^2\right)\phi =  ikU_{yy}\phi$		& \multicolumn{2}{c}{Equation \eqref{Orr-somerfield}} \\
 Critical parameters	& none						& $\Rey \propto \Ndg^{1/2}$ 	& $\Rey \propto \Ndg$ \\
 Most unstable $k$ as $\delta \to 0$	& $\propto \delta^{-1}$		& $\propto \delta^{-1}$	& $O(1)$ \\
 Mode localized?	& yes, near $y=0$				& ~~~~yes, near $y=\hg$~~~~			& no
 \end{tabular}
 \caption{Comparison between Kelvin-Helmholtz instability and the two modes resulting from solution of \ref{Orr-somerfield}.}
 \label{tab:comparison}
\end{table*}
To better understand the mechanism of waving and the origin of the threshold, we consider the behavior of the instability as a function of the vegetation drag characterized by $\Ndg$.
The fastest growing wavenumber first increases proportional to $H/\delta$, but at a critical vegetation density discontinuously jumps and remains $O(1)$ (see fig.~\ref{Re_vs_delta}). 
To aid in explaining the behavior, we show heat maps of Re$(\sigma)$ as a function of $\Rey$ and $k$, for different $\hg/H$ and $\Ndg$ in fig.~\ref{K_Re_sigma_set3}. 
The lowest $\Rey$ on the neutral curve sets the threshold. 
We observe that as $\Ndg$ increases, the unstable region splits into two. We refer to the unstable region with the higher $k$ as ``Mode 1'', and the one with the lower $k$ as ``Mode 2''. 
The unstable region for Mode 1, depending on $\hg/H$ either recedes to higher $\Rey$ or shrinks to zero size, as the vegetation density increases, causing the most unstable mode to transition discontinuously.

The distinct asymptotic behavior of the two modes as $\Ndg \gg 1$ allows us to understand the mechanism of the instability. 
Mode 1 asymptotically localizes to the boundary layer near the top of the vegetation, and exhibits an asymptotic behavior with $k \sim O(H/\delta)$, and $\Rey \sim (H/\delta)$ (or $\Rey \propto {\Ndg}^{1/2}$) at the threshold. 
This limit can be understood through asymptotically estimating the sizes of the terms in eq.~\eqref{Orr-somerfield} using $D\sim \delta^{-1}$, $\sigma \sim \delta^{-1}$, and $U\sim \delta$; the magnitude of the advection term is $\Rey/\delta^3 \sim \Rey^2 \Ndg $, and the viscous term (or the vegetation drag term, which is found to scale identically) in the boundary layer is $\delta^{-4} \sim (\Rey \Ndg)^{4/3}$. 
The terms balance when $\Rey \sim H/\delta$ (or $\Rey \sim {\Ndg}^{1/2}$) leading to a simplification of eq.~\eqref{Orr-somerfield} as
\begin{align}
(D^2-k^2)^2 \phi- 2 \Rey \Ndg D( SUD\phi) = \sigma \Rey (D^2-k^2)\phi,
\label{eqn:mode1asymp}
\end{align} 
in a region of thickness O($\delta$) near $y=\hg$.
Since the relative magnitude of the terms on the r.h.s. to those on the l.h.s. of this equation is $\Rey/\Ndg^{1/2}$, we expect identical asymptotic behavior for fixed $\Rey/\Ndg^{1/2}$. Therefore the threshold obtained for Mode 1 is $\Rey \propto \Ndg^{1/2}$ (or $\Rey \propto H/\delta$) explaining the numerically observed asymptote (see fig.~\ref{Re_vs_delta}). 
This analysis also concludes that the mode structure is self-similar over the length scale $\delta$ for fixed $\Rey/\Ndg^{1/2}$; the verification of this expectation is shown in fig.~\ref{Asymptotic_mode}, supporting this argument.

On the other hand, the threshold condition for Mode 2 is numerically observed to be $\Rey \propto ({\delta}/{H})^{-3/2}$ (or $\Rey \propto \Ndg$) for $k\sim O(1)$, shown in fig.~\ref{Re_vs_delta}, which can be understood by an asymptotic scaling analysis of eq.~\eqref{Orr-somerfield} by assuming $\Rey \gg 1$ but fixed $\Rey/\Ndg \sim O(1)$.
Note that for large $\Ndg$, the non-dimensional steady flow velocity inside the vegetation is $U_g \sim (\Rey \Ndg)^{-1/2}$. 
This estimate simplifies eq.~\eqref{Orr-somerfield} to 
\begin{align}
\begin{split}
\sigma\left(D^2-k^2\right)\phi = -2{(\Ndg/\Rey)^{1/2}}D^2\phi,  \quad &\text{ for $y<\hg$}  \\
\left(\sigma+ikU\right) \left(D^2-k^2\right)\phi =  ikU_{yy}\phi, \quad &\text{ for $y>\hg$}.
\end{split}
\label{eqn:mode2asymp}
\end{align}
\begin{figure}
\includegraphics[]{Asymptotic_noshear}
\caption{Plot of mode shape $|\phi|$ for different shear layer thickness in limit of small shear layer thickness for representative values of vegetation density. 
Mode 1 is shown in solid and Mode 2 is shown in dashed. The parameters Mode 2 shapes are chosen such that $\Rey \gg 1$, $\Ndg \gg 1$ (specified in terms of $\delta/H$) but $\Rey/\Ndg = O(1)$. 
The mode shapes approach each other for these small values of $\delta/H$ indicating that the large vegetation density asymptote is reached. Mode 1 shapes appear self-similar in shape as $\delta\to 0$, 
and are compared to each other on a rescaled coordinate in the inset. 
Inset shows $|\phi|$ for Mode 1 as a function of $(y-\hg)/\delta$. The modes approach a universal shape, indicating that an asymptotic limit has been reached. 
The limit is not yet reached for the case $\delta/H = 0.005$ due to the influence of bottom boundary; the vegetation height in this case is comparable to the boundary layer thickness.}
\label{Asymptotic_mode}
\end{figure}
Since $\Rey/\Ndg$ is the only remaining parameter, the mode shape would converge in the aforementioned limit, in agreement with our numerical results for Mode 2 shown in fig.~\ref{Asymptotic_mode}. 
We interpret Mode 2 as the instability of an inviscid flow, with the vegetation modeled by a continuum drag field, and for which the boundary layer near the top of the vegetation plays no role. The only remaining parameter $\Rey/\Ndg$ 
sets the threshold, leading to the asymptotic behavior $\Rey \propto \Ndg$ (or $\Rey \sim ({\delta}/{H})^{-3/2}$).
%On the other hand, Mode 1 asymptotically localizes to the boundary layer near the grass tip, and exhibits a different asymptotic behavior with $k \sim O(H/\delta)$, and $\Rey \sim (H/\delta)$ (or $\Rey \propto {\Ndg}^{1/2}$) at the threshold.
